Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models

### 2025-04-11 brms_meeg.tex
new sections in the intro

\documentclass[
  man,
  floatsintext,
  longtable,
  a4paper,
  nolmodern,
  notxfonts,
  notimes,
  colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{apa7}

@@

@@

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}
\usepackage{anyfontsize}
\usepackage{multirow}
\usepackage{mathtools}
\usepackage{booktabs, caption, longtable, colortbl, array}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

The conventional approach involves mass-univariate statistical testing
through time and/or space followed by some form or correction for
multiple comparisons with the goal of maintaining the familywise error
rate (FWER) or false discovery rate (FDR) at the nominal level (e.g.,
5\%). Cluster-based inference is the most common way of achieving this
sort of error control in the M/EEG literature, being the recommended
approach in several software programs (e.g., \texttt{EEGlab} or
\texttt{MNE-Python}, \citeproc{ref-delorme2004}{Delorme \& Makeig,
2004}; \citeproc{ref-gramfort2013}{Gramfort, 2013}). While effective for
controlling error rates, cluster-based inference comes with a
significant limitation: by shifting the focus of inference from
individual time points to clusters, it prevents the ability to draw
precise conclusions about the onset or offset of observed effects
(\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007};
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019}). As
pointed out by Maris \& Oostenveld (\citeproc{ref-maris2007}{2007});
``there is a conflict between this interest in localized effects and our
choice for a global null hypothesis: by controlling the FA {[}false
alarm{]} rate under this global null hypothesis, one cannot quantify the
uncertainty in the spatiotemporal localization of the effect''. Even
worse, Rosenblatt et al. (\citeproc{ref-rosenblatt2018}{2018}) note that
cluster-based inference suffers from low spatial resolution: ``Since
discovering a cluster means that `there exists at least one voxel with
an evoked response in the cluster', and not that `all the voxels in the
cluster have an evoked response', it follows that the larger the
detected cluster, the less information we have on the location of the
activation.'' As a consequence, cluster-based inference is expected to
perform poorly for localising the onset of M/EEG effects; a property
that was later demonstrated in simulations studies (e.g.,
\citeproc{ref-rousselet_using_2025}{Rousselet, 2025};
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019}).

To overcome the limitations of cluster-based inference, we introduce a
novel \emph{model-based} approach for precisely localising M/EEG effects
in time, space, and other dimensions. The proposed approach, based on
Bayesian generalised additive multilevel models, allows quantifying the
posterior probability of effects being above chance at the level of
timesteps, sensors, voxels, or other dimensions, while naturally taking
into account spatiotemporal dependencies present in M/EEG timeseries. We
compare the performance of the proposed approach to well-established
alternative methods using both simulated and actual M/EEG data and show
that it significantly outperforms alternative methods in estimating the
onset and offset of M/EEG effects.

\subsection{Cluster-based inference}\label{cluster-based-inference}

The issue with multiple comparisons\ldots{} Different methods exist to
control the family-wiser error rate (FWER), defined as the type-1 error
rate (false positive) over an ensemble (family) of tests\ldots{} for
instance the Bonferroni correction (\citeproc{ref-dunn1961}{Dunn,
1961}), however this method is generally overconservative as it assumes
statistical independence of tests, an assumptions that is clearly
violated in the context of M/EEG timeseries with spatiotemporal
dependencies\ldots{} or the false discovery rate (FDR), defined as the
proportion of false positive among positive tests (e.g.,
\citeproc{ref-benjamini1995}{Benjamini \& Hochberg, 1995};
\citeproc{ref-benjamini2001}{Benjamini \& Yekutieli, 2001})\ldots{}

On popular technique to account for spatiotemporal dependencies while
controlling the FWER is cluster-based inference. Description of
cluster-based approaches (see \citeproc{ref-sassenhagen2019}{Sassenhagen
\& Draschkow, 2019})\ldots{} A typical cluster-based inference consists
of two successive steps. First, clusters are defined as sets of
contiguous voxels, channels, and/or timesteps, whose intensity/activity
exceeds some predefined threshold. Clusters are then characterised by
their height (i.e., maximal value), extent (number of constituent
elements), or some combination of both, such at its ``mass'', for
instance by summing the statistics within a cluster, an approach
refereed to as ``cluster mass'' (\citeproc{ref-maris2007}{Maris \&
Oostenveld, 2007}; \citeproc{ref-pernet2015}{Pernet et al., 2015}).
Then, the null hypothesis is tested by assessing the probability.. As
different cluster-forming thresholds lead to clusters with different
spatial or temporal extent, this initial threshold modulates the
sensitivity of the subsequent permutation test. The threshold-free
cluster enhancement method (TFCE) was introduced by S. Smith \& Nichols
(\citeproc{ref-smith2009}{2009}) to overcome this choice of an arbitrary
threshold.

In brief, the TFCE method works as follows. Instead of picking an
arbitrary cluster-forming threshold (e.g., \(t=2\)), we try all (or
many) possible thresholds in a given range and check whether a given
timestep/voxel belongs to a significant cluster under any of the set of
thresholds. Then, instead of using cluster mass (e.g., the sum of
squared t-values within the cluster), we use a weighted average between
the cluster extend (\(e\), how broad is the cluster, that is, how many
connected samples it contains) and the cluster height (\(h\), how high
is the cluster, that is, how large is the test statistic). The TFCE
score at each timestep/voxel \(t\) is given by:

\[
\text{TFCE(t)} = \int_{h=h_{0}}^{h_{t}} e(h)^{E} h^{H} \mathrm{d}h
\]

where \(h_{0}\) is typically \(0\) and parameters \(E\) and \(H\) are
set a priori (typically to 0.5 and 2, respectively) and control the
influence of the extend and height on the TFCE. Note that in practise,
this intergral is approximated by a sum over small \(h\) increments.
Then, p-value for timestep/voxel \(t\) is computed by comparing it TFCE
with the null distribution of TFCE values. For each permuted signal, we
keep the maximal value over the whole signal for the null distribution
of the TFCE. The TFCE combined with permutation (assuming a large enough
number of permutations) has been shown to provide accurate type 1 FWER
(e.g., \citeproc{ref-pernet2015}{Pernet et al., 2015}). However,
previous simulation work showed that cluster-based methods (including
the TFCE method) perform poorly in localising the onset of M/EEG effects
(e.g., \citeproc{ref-rousselet_using_2025}{Rousselet, 2025};
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019}).

To sum up this section, cluster-based inference main limitations is that
it provides inference at the cluster level only, not allowing inference
at the level of timesteps, sensors, etc. As a consequence, it does not
allow inferring the localisation of effects (for more details on
cluster-based inference, see for instance
\citeproc{ref-maris2011}{Maris, 2011}; \citeproc{ref-maris2007}{Maris \&
Oostenveld, 2007}; \citeproc{ref-sassenhagen2019}{Sassenhagen \&
Draschkow, 2019}). In the following, we briefly review previous M/EEG
modelling work. Then, we provide a brief introduction to generalised
additive models (GAMs) and Bayesian generalised additive multilevel
models (BGAMMs) to illustrate how these models can be used to precisely
localise the onset and offset of M/EEG effects.

@@

From Dimigen \& Ehinger (\citeproc{ref-dimigen2021}{2021}): Recently,
spline regression has been applied to ERPs (e.g.,
\citeproc{ref-hendrix2017}{Hendrix et al., 2017};
\citeproc{ref-kryuchkova2012}{Kryuchkova et al., 2012})\ldots{} GAMMs
for EEG data (\citeproc{ref-abugaber2023}{Abugaber et al., 2023};
\citeproc{ref-meulman2015}{Meulman et al., 2015},
\citeproc{ref-meulman2023}{2023})\ldots{}

@@

where \(\beta_{jk}\) is the weight (coefficient) associated with the
\(k\)th basis function \(b_{jk}()\) evaluated at the covariate value
\(x_{ij}\) for the \(j\)th smooth function \(f_{j}\). To clarify the
terminology at this stage: \emph{splines} are functions composed of
simpler functions. These simpler functions are basis functions (e.g.,
cubic polynomial, thin-plate) and the set of basis functions is a
\emph{basis}. Each basis function gets its coefficient and the resultant
spline is the sum of these weighted basis functions
(Figure~\ref{fig-intro-gam}). Splines coefficients are penalised
(usually through the squared of the smooth functions' second derivative)
in a way that can be interpreted, in Bayesian terms, as a prior on the
``wiggliness'' of the function. In other words, more complex (wiggly)
basis function are penalised.

@@

However, the previous model only included constant (fixed) effects, thus
not properly accounting for between-participant variability. We next fit
a multilevel version of the BGAM (BGAMM, see
\citeproc{ref-nalborczyk2019}{Nalborczyk et al., 2019}). Although it is
possible to fit a BGAMM at the single-trial level, we present a
computationally-lighter version of the model that is fitted directly on
by-participant summary statistics (mean and SD), similar to what is done
in meta-analysis.

@@

\CommentTok{\# fitting the BGAMM}
\NormalTok{meta\_gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# using by{-}participant SD of ERPs across trials}
\NormalTok{    eeg\_mean }\SpecialCharTok{|} \FunctionTok{se}\NormalTok{(eeg\_sd) }\SpecialCharTok{\textasciitilde{}}
\NormalTok{        condition }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(time, }\AttributeTok{bs =} \StringTok{"cr"}\NormalTok{, }\AttributeTok{k =} \DecValTok{20}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition) }\SpecialCharTok{+}
\NormalTok{        (}\DecValTok{1} \SpecialCharTok{|}\NormalTok{ participant),}
    \AttributeTok{data =}\NormalTok{ summary\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{warmup =} \DecValTok{2000}\NormalTok{,}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/meta\_gam.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

@@

We then compute the posterior probability of the slope for
\texttt{condition} being above \(0\) (Figure~\ref{fig-post-prob-ratio},
left). From this quantity, we then compute the ratio of posterior
probabilities (i.e., \(p/(1-p)\)) and visualise the timecourse of this
ratio superimposed with the conventional thresholds on evidence ratios
(Figure~\ref{fig-post-prob-ratio}, right). Note that a ratio of 10 means
that the probability of the difference being above 0 is 10 times higher
than the probability of the difference not being above 0, given the
data, the priors, and other model's assumptions. Thresholding the
posterior probability ratio thus provides a model-based approach for
estimating the onset and offset of M/EEG effects.

@@

We then compared the ability of the BGAMM to correctly estimate the
onset and offset of the ERP difference to other widely-used methods.
First, we conducted mass-univariate t-tests (thus treating each timestep
independently) and identified the onset and offset of the ERP difference
as the first and last values crossing an arbitrary significance
threshold (\(\alpha = 0.05\)). We then followed the same approach but
after applying different forms of multiplicity correction to the
p-values. We compared two methods that control the false discovery rate
(FDR) (i.e., BH95, \citeproc{ref-benjamini1995}{Benjamini \& Hochberg,
1995}; and BY01, \citeproc{ref-benjamini2001}{Benjamini \& Yekutieli,
2001}), one method that controls the family-wise error rate (FWER)
(i.e., Holm--Bonferroni method, \citeproc{ref-holm1979}{Holm, 1979}),
and two cluster-based permutation methods (permutation with a single
cluster-forming threshold and threshold-free cluster enhancement, TFCE,
\citeproc{ref-smith2009}{S. Smith \& Nichols, 2009}). The BH95, BY01,
and Holm corrections were applied to the p-values using the
\texttt{p.adjust()} function in \texttt{R}. The cluster-based inference
was implemented using a cluster-sum statistic of squared t-values, as
implemented in \texttt{MNE-Python}
(\citeproc{ref-gramfort2013}{Gramfort, 2013}), called via the \texttt{R}
package \texttt{reticulate} v 1.42.0 (\citeproc{ref-reticulate}{Ushey et
al., 2024}). We also compared these estimates to the onset and offset as
estimated using the binary segmentation algorithm, as implemented in the
\texttt{R} package \texttt{changepoint} v 2.3
(\citeproc{ref-changepoint}{Killick et al., 2022}), and applied directly
to the squared t-values (as in
\citeproc{ref-rousselet_using_2025}{Rousselet, 2025}).
Figure~\ref{fig-corrections} illustrates the onsets and offsets
estimated by each method on a single simulated dataset and shows that
all methods systematically overestimate the true onset and underestimate
the true offset.

@@

To assess the accuracy of group-level onset estimation, the various
methods were compared using the bias (i.e., median(estimated-true)),
median absolute error (MAE), root mean square error (RMSE), variance,
and median absolute deviation (MAD) of onset/offset estimates computed
on 1,000 simulated datasets. As in Rousselet
(\citeproc{ref-rousselet_using_2025}{2025}), each participant was
assigned a random onset between 150 and 170ms.

@@

To complement the simulation study, we evaluated the performance of the
various methods on actual MEG data (decoding results from
\citeproc{ref-nalborczyk:inprep}{Nalborczyk et al., in preparation}). In
this study, we conducted time-resolved multivariate pattern analysis
(MVPA, also known as decoding) of MEG data during reading tasks. As a
result, we obtain a timecourse of decoding performance (ROC AUC),
bounded between 0 and 1, for each participant (for a total of 32
participants). Next, we wanted to \emph{test} whether the group-level
average decoding accuracy is above chance (i.e., 0.5) at each timestep
(Figure~\ref{fig-decoding-data}). To achieve this, we fitted a BGAM as
introduced previously, but we replaced the \(\mathrm{Normal}\)
likelihood function by a \(\mathrm{Beta}\) one to account for the
bounded nature of AUC values (between 0 and 1) (for a tutorial on Beta
regression, see \citeproc{ref-coretta2025}{Coretta \& Bürkner, 2025}).

>>>>

\documentclass[
  doc,
  floatsintext,
  longtable,
  a4paper,
  nolmodern,
  notxfonts,
  notimes,
  donotrepeattitle,
  colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{apa7}

@@

\geometry{inner=1in, outer=1in}
\fancyhfoffset[LE,RO]{0cm}

@@

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}
\usepackage{anyfontsize}
\usepackage{multirow}
\usepackage{mathtools}
\usepackage{booktabs, caption, longtable, colortbl, array}
\usepackage{setspace}
\onehalfspacing
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

The conventional approach involves mass-univariate statistical testing
through time and/or space followed by some form or correction for
multiple comparisons with the goal of maintaining the familywise error
rate (FWER) or the false discovery rate (FDR) at the nominal level
(e.g., 5\%). Cluster-based inference is the most common way of achieving
this sort of error control in the M/EEG literature, being the
recommended approach in several software programs (e.g.,
\texttt{EEGlab}, \citeproc{ref-delorme2004}{Delorme \& Makeig, 2004};
\texttt{MNE-Python}, \citeproc{ref-gramfort2013}{Gramfort, 2013}). While
effective for controlling error rates, cluster-based inference comes
with a significant limitation: by shifting the focus of inference from
individual datapoints (e.g., timesteps, sensors, voxels) to clusters, it
prevents the ability to draw precise conclusions about the
spatiotemporal localisation of such effects
(\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007};
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019}). As
pointed out by Maris \& Oostenveld (\citeproc{ref-maris2007}{2007}):
``there is a conflict between this interest in localized effects and our
choice for a global null hypothesis: by controlling the FA {[}false
alarm{]} rate under this global null hypothesis, one cannot quantify the
uncertainty in the spatiotemporal localization of the effect''. Even
worse, Rosenblatt et al. (\citeproc{ref-rosenblatt2018}{2018}) note that
cluster-based inference suffers from low spatial resolution: ``Since
discovering a cluster means that `there exists at least one voxel with
an evoked response in the cluster', and not that `all the voxels in the
cluster have an evoked response', it follows that the larger the
detected cluster, the less information we have on the location of the
activation.'' As a consequence, cluster-based inference is expected to
perform poorly for localising the onset of M/EEG effects; a property
that was later demonstrated in simulations studies (e.g.,
\citeproc{ref-rousselet_using_2025}{Rousselet, 2025};
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019}).

To overcome the limitations of cluster-based inference, we introduce a
novel \emph{model-based} approach for precisely localising M/EEG effects
in time, space, and other dimensions. The proposed approach, based on
Bayesian generalised additive multilevel models, allows quantifying the
posterior probability of effects being above chance at the level of
timesteps, sensors, voxels, etc, while naturally taking into account
spatiotemporal dependencies present in M/EEG data. We compare the
performance of the proposed approach to well-established alternative
methods using both simulated and actual M/EEG data and show that it
significantly outperforms alternative methods in estimating the onset
and offset of M/EEG effects.

\subsection{Statistical errors and cluster-based
inference}\label{statistical-errors-and-cluster-based-inference}

The issues with multiple comparisons represent a common and
well-recognised danger in neuroimaging and M/EEG research where the
collected data allows for a multitude of potential hypothesis tests and
is characterised by complex structures of spatiotemporal dependencies.
The probability of obtaining at least one false positive in an ensemble
(family) of \(m\) tests (i.e., the FWER) is computed as
\(1-\left(1-\alpha\right)^{m}\) (for \(m=10\) tests and \(\alpha=0.05\),
it is approximately equal to \(0.4\)). Different methods exist to
control the FWER (i.e., bring it back to \(\alpha\)). Most methods apply
a simple correction to series of ``raw'' \(p\)-values issued from
univariate statistical tests (e.g., t-tests). For instance, the
Bonferroni correction (\citeproc{ref-dunn1961}{Dunn, 1961}) consists in
setting the significance threshold to \(\alpha/m\), or equivalently,
multiplying the ``raw'' \(p\)-values by \(m\) and using the standard
\(\alpha\) significance threshold. This method is generally
overconservative (i.e., under-powered) as it assumes statistical
independence of the tests, an assumption that is clearly violated in the
context of M/EEG timeseries characterised by massive spatiotemporal
dependencies. Some alternative methods aims at controlling the FDR,
defined as the proportion of false positive \emph{among positive tests}
(e.g., \citeproc{ref-benjamini1995}{Benjamini \& Hochberg, 1995};
\citeproc{ref-benjamini2001}{Benjamini \& Yekutieli, 2001}). However, a
major limitation of both types of corrections is that they do not take
into account the spatial and temporal information contained in M/EEG
data.

A popular technique to account for spatiotemporal dependencies while
controlling the FWER is cluster-based inference
(\citeproc{ref-bullmore1999}{Bullmore et al., 1999};
\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007}). A typical
cluster-based inference consists of two successive steps (for more
details on cluster-based inference, see for instance
\citeproc{ref-frossard2022}{Frossard \& Renaud, 2022};
\citeproc{ref-maris2011}{Maris, 2011}; \citeproc{ref-maris2007}{Maris \&
Oostenveld, 2007}; \citeproc{ref-sassenhagen2019}{Sassenhagen \&
Draschkow, 2019}). First, clusters are defined as sets of contiguous
timesteps, sensors, voxels, etc, whose activity, summarised by some test
statistic (e.g., a \(t\)-value), exceeds a predefined threshold (e.g.,
the 95th percentile of the parametric null distribution). Clusters are
then characterised by their height (i.e., maximal value), extent (number
of constituent elements), or some combination of both, for instance by
summing the statistics within a cluster, an approach referred to as
``cluster mass'' (\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007};
\citeproc{ref-pernet2015}{Pernet et al., 2015}). Then, the null
hypothesis is tested by computing a \(p\)-value for each identified
cluster by comparing its mass with the null distribution of cluster
masses (obtained via permutation). As alluded previously, a significant
cluster is a cluster which contains \emph{at least one} significant
time-point. As such, it would be incorrect to conclude, for instance,
that the timestep of a significant cluster is the first moment at which
some conditions differ (\citeproc{ref-frossard2022}{Frossard \& Renaud,
2022}; \citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019}).
In other words, because the inference is performed at the second step
(i.e., once clusters have been formed), it prevents any conclusion to be
made about individual datapoints (e.g., timesteps, sensors, etc).

As different cluster-forming thresholds lead to clusters with different
spatial or temporal extent, this initial threshold modulates the
sensitivity of the subsequent permutation test. The threshold-free
cluster enhancement (TFCE) method was introduced by S. Smith \& Nichols
(\citeproc{ref-smith2009}{2009}) to overcome this choice of an arbitrary
threshold. In brief, the TFCE method works as follows. Instead of
picking an arbitrary cluster-forming threshold (e.g., \(t=2\)), the
methods consist in trying all (or many) possible thresholds in a given
range and checking whether a given datapoint (e.g., timestep, sensor,
voxel) belongs to a significant cluster under any of the set of
thresholds. Then, instead of using cluster mass, one uses a weighted
average between the cluster extend (\(e\), how broad is the cluster,
that is, how many connected samples it contains) and the cluster height
(\(h\), how high is the cluster, that is, how large is the test
statistic). The TFCE score at each timestep \(t\) is given by:

\[
\text{TFCE(t)} = \int_{h=h_{0}}^{h=h_{t}} e(h)^{E} h^{H} \mathrm{d}h
\]

where \(h_{0}\) is typically \(0\) and parameters \(E\) and \(H\) are
set a priori (typically to 0.5 and 2, respectively) and control the
influence of the extend and height on the TFCE. Note that in practise,
this integral is approximated by a sum over small \(h\) increments.
Then, a p-value for each timestep \(t\) is computed by comparing its
TFCE with the null distribution of TFCE values (obtained via
permutation). For each permuted signal, we keep the maximal value over
the whole signal for the null distribution of the TFCE. The TFCE
combined with permutation (assuming a large enough number of
permutations) has been shown to provide accurate FWER control (e.g.,
\citeproc{ref-pernet2015}{Pernet et al., 2015}). However, further
simulation work showed that cluster-based methods (including TFCE)
perform poorly in localising the onset of M/EEG effects (e.g.,
\citeproc{ref-rousselet_using_2025}{Rousselet, 2025};
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019}).

To sum up, the main limitation of cluster-based inference is that it
allows for inference at the cluster level only, not allowing inference
at the level of timesteps, sensors, etc. As a consequence, it does not
allow inferring the precise spatial and temporal localisation of
effects. In the following, we briefly review previous M/EEG modelling
work. Then, we provide a short introduction to generalised additive
models (GAMs) and Bayesian generalised additive multilevel models
(BGAMMs) to illustrate how these models can be used to precisely
localise the onset and offset of M/EEG effects.

@@

From Dimigen \& Ehinger (\citeproc{ref-dimigen2021}{2021}): Recently,
spline regression has been applied to ERPs (e.g.,
\citeproc{ref-hendrix2017}{Hendrix et al., 2017};
\citeproc{ref-kryuchkova2012}{Kryuchkova et al., 2012})\ldots{} GAMMs
for EEG data (\citeproc{ref-abugaber2023}{Abugaber et al., 2023};
\citeproc{ref-meulman2015}{Meulman et al., 2015},
\citeproc{ref-meulman2023}{2023})\ldots{} See also the UNFOLD toolbox
(\citeproc{ref-ehinger_unfold_2019}{Ehinger \& Dimigen, 2019})\ldots{}

@@

where \(\beta_{jk}\) is the weight (coefficient) associated with the
\(k\)th basis function \(b_{jk}()\) evaluated at the covariate value
\(x_{ij}\) for the \(j\)th smooth function \(f_{j}\). To clarify the
terminology at this stage: \emph{splines} are functions composed of
simpler functions. These simpler functions are basis functions (e.g.,
cubic polynomial, thin-plate) and the set of basis functions is a
\emph{basis}. Each basis function gets its coefficient and the resultant
spline is the sum of these weighted basis functions
(Figure~\ref{fig-intro-gam}). Splines coefficients are penalised
(usually through the squared of the smooth functions' second derivative)
in a way that can be interpreted, in Bayesian terms, as a prior on the
``wiggliness'' of the function. In other words, more complex (wiggly)
basis functions are penalised.

@@

However, the previous model only included constant (fixed) effects, thus
not properly accounting for between-participant variability. We next fit
a multilevel version of the BGAM (BGAMM, for an introduction to Bayesian
multilevel models in \texttt{brms}, see
\citeproc{ref-nalborczyk2019}{Nalborczyk et al., 2019}). Although it is
possible to fit a BGAMM at the single-trial level, we present a
computationally-lighter version of the model that is fitted directly on
by-participant summary statistics (mean and SD), similar to what is done
in meta-analysis.

@@

\CommentTok{\# fitting the BGAMM}
\NormalTok{meta\_gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# using by{-}participant SD of ERPs across trials}
\NormalTok{    eeg\_mean }\SpecialCharTok{|} \FunctionTok{se}\NormalTok{(eeg\_sd) }\SpecialCharTok{\textasciitilde{}}
\NormalTok{        condition }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(time, }\AttributeTok{bs =} \StringTok{"cr"}\NormalTok{, }\AttributeTok{k =} \DecValTok{20}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition) }\SpecialCharTok{+}
\NormalTok{        (}\DecValTok{1} \SpecialCharTok{|}\NormalTok{ participant),}
    \AttributeTok{data =}\NormalTok{ summary\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{warmup =} \DecValTok{2000}\NormalTok{,}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/meta\_gam.rds"}
\NormalTok{    )}

\CommentTok{\# fitting the BGAMM}
\CommentTok{\# meta\_gamm \textless{}{-} brm(}
\CommentTok{\#     \# using by{-}participant SD of ERPs across trials}
\CommentTok{\#     eeg\_mean | se(eeg\_sd) \textasciitilde{} condition +}
\CommentTok{\#         s(participant, bs = "re") +}
\CommentTok{\#         s(time, participant, bs = "fs", by = condition, m = 1),}
\CommentTok{\#     data = summary\_df,}
\CommentTok{\#     family = gaussian(),}
\CommentTok{\#     warmup = 2000,}
\CommentTok{\#     iter = 5000,}
\CommentTok{\#     chains = 8,}
\CommentTok{\#     cores = 8,}
\CommentTok{\#     control = list(adapt\_delta = 0.95),}
\CommentTok{\#     file = "models/meta\_gamm.rds"}
\CommentTok{\#     )}
\end{Highlighting}
\end{Shaded}

@@

We then compute the posterior probability of the slope for
\texttt{condition} being above \(0\) (Figure~\ref{fig-post-prob-ratio},
left). From this quantity, we then compute the ratio of posterior
probabilities (i.e., \(p/(1-p)\)) and visualise the timecourse of this
ratio superimposed with the conventional thresholds on evidence ratios
(Figure~\ref{fig-post-prob-ratio}, right). Note that a ratio of 10 means
that the probability of the difference being above 0 is 10 times higher
than the probability of the difference not being above 0, given the
data, the priors, and other model's assumptions. Thresholding the
posterior probability ratio thus provides a model-based approach for
estimating the onset and offset of M/EEG effects. An important advantage
is that the proposed approach can be extended to virtually any model
structure. Moreover, the approach provides both group-level and
individual-level onset and offset estimates of M/EEG effects.

@@

We then compared the ability of the BGAMM to accurately estimate the
onset and offset of the ERP difference to other widely-used methods.
First, we conducted mass-univariate t-tests (thus treating each timestep
independently) and identified the onset and offset of the ERP difference
as the first and last values crossing an arbitrary significance
threshold (\(\alpha = 0.05\)). We then followed the same approach but
after applying different forms of multiplicity correction to the
\(p\)-values. We compared two methods that control the FDR (i.e.,
\texttt{BH95}, \citeproc{ref-benjamini1995}{Benjamini \& Hochberg,
1995}; and \texttt{BY01}, \citeproc{ref-benjamini2001}{Benjamini \&
Yekutieli, 2001}), one method that controls the FWER (i.e.,
Holm--Bonferroni method, \citeproc{ref-holm1979}{Holm, 1979}), and two
cluster-based permutation methods (permutation with a single
cluster-forming threshold and threshold-free cluster enhancement,
\texttt{TFCE}, \citeproc{ref-smith2009}{S. Smith \& Nichols, 2009}). The
\texttt{BH95}, \texttt{BY01}, and \texttt{Holm} corrections were applied
to the p-values using the \texttt{p.adjust()} function in \texttt{R}.
The cluster-based inference was implemented using a cluster-sum
statistic of squared \(t\)-values, as implemented in \texttt{MNE-Python}
(\citeproc{ref-gramfort2013}{Gramfort, 2013}), called via the \texttt{R}
package \texttt{reticulate} v 1.42.0 (\citeproc{ref-reticulate}{Ushey et
al., 2024}). We also compared these estimates to the onset and offset as
estimated using the binary segmentation algorithm, as implemented in the
\texttt{R} package \texttt{changepoint} v 2.3
(\citeproc{ref-changepoint}{Killick et al., 2022}), and applied directly
to the squared \(t\)-values (as in
\citeproc{ref-rousselet_using_2025}{Rousselet, 2025}).
Figure~\ref{fig-corrections} illustrates the onsets and offsets
estimated by each method on a single simulated dataset and shows that
all methods systematically overestimate the true onset and underestimate
the true offset.

@@

To assess the accuracy of group-level onset estimation, the various
methods were compared using the bias (i.e., median(estimated-true)),
median absolute error (MAE), root mean square error (RMSE), variance,
and median absolute deviation (MAD) of onset/offset estimates computed
on 1,000 simulated datasets. As in Rousselet
(\citeproc{ref-rousselet_using_2025}{2025}), each participant was
assigned a random onset between 150 and 170ms. Whereas the present
article focuses on one-dimensional signas (e.g., one M/EEG channel), we
provice a 2D application in Section~\ref{sec-2D}.

@@

To complement the simulation study, we evaluated the performance of the
various methods on actual MEG data (decoding results from
\citeproc{ref-nalborczyk:inprep}{Nalborczyk et al., in preparation}). In
this study, we conducted time-resolved multivariate pattern analysis
(MVPA, also known as decoding) of MEG data during reading tasks. As a
result, we obtain a timecourse of decoding performance (ROC AUC),
bounded between 0 and 1, for each participant (\(N=32\)). Next, we
wanted to \emph{test} whether the group-level average decoding accuracy
is above chance (i.e., 0.5) at each timestep
(Figure~\ref{fig-decoding-data}). To achieve this, we fitted a BGAM as
introduced previously, but we replaced the \(\mathrm{Normal}\)
likelihood function by a \(\mathrm{Beta}\) one to account for the
bounded nature of AUC values (between 0 and 1) (for a tutorial on Beta
regression, see \citeproc{ref-coretta2025}{Coretta \& Bürkner, 2025}).

Note that although we chose a basis dimension of \(k=50\), which seems
appropriate for the present data, this choice should be adapted
according to the properties of the modelled data (e.g., signal-to-noise
ratio, prior low-pass filtering, sampling rate) and should be assessed
by the usual model checking tools (e.g., posterior predictive checks,
see also Section~\ref{sec-basis}). To better distinguish signal from
noise, we also defined a region of practical equivalence (ROPE,
\citeproc{ref-kruschke2017}{Kruschke \& Liddell, 2017}), defined as the
chance level plus the standard deviation of the (group-level average)
decoding performance during the baseline period.
